Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-25 joint_inf.tex
sketch of proof of inference limit ==> limit inference

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\end{equation}
%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data? Compare https://paperpile.com/shared/HohiVH in which they are rather different.
%das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

>>>>

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Use $\ell_{\tau^*,t^*}(\tau, t)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.

%EM Could we give an extra sentence justifying that the limit of the inferential process will be the same as the inferences applied to the limiting data? Compare https://paperpile.com/shared/HohiVH in which they are rather different.
%das: wow, good catch! This is not at all obvious to me it turns out :/ I guess I'll have to think harder on it
%das: update: I think I've got it, but I need to write it up a little more explicitly, and probably in the appendix. I think this approach maybe doesn't apply to your reference? But I'd like to work through it and see if that's the case, as a sanity check.
Let
\[
\hat{t} = \arg\max_{t\in\mathcal{T}} \ \ell_{\tau^*,t^*}(\tau, t)
\]
and
\[
m_t(\alignmentColumn) = \sum_{j=1}^\nSiteSplits 1\{\patternToSplit(\alignmentColumn)=\siteSplit_j\} \cdot [\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\]
so that for a fixed $\tau$
\[
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) = \frac{1}{n} \sum_{i=1}^\nCols m_t(\alignmentColumn_i)
\]
and
\[
\ell_{\tau^*,t^*}(\tau, t) = E[m_t].
\]
To show $\hat{t}_\nCols \rightarrow \hat{t}$, we use Wald's consistency proof \citep[p. 48, Theorem 5.14 of ][]{van1998asymptotic}, which requires the following:
\begin{itemize}
\item $\mathcal{T}$ is compact;
\item $t \mapsto m_t(\alignmentColumn)$ is upper-semicontinuous for almost all $\alignmentColumn$;
\item $\alignmentColumn \mapsto \sup_t m_t(\alignmentColumn)$ is measurable;
\item $E[\sup_t m_t] < \infty$.
\end{itemize}
Point one is obviously true.
Point two is true as well: for a fixed ancestral state split $t \mapsto m_t(\alignmentColumn)$ is continuous for all $\alignmentColumn$.
Discontinuities only occur when, if we move about in $\mathcal{T}$, a different ancestral state split becomes more likely.
Maximizing over ancestral state splits ensures the upper-semicontinuous property of this mapping.
For point three, we can think of $\sup_t m_t$ as fixing a maximal ancestral state split.
Since $\alignmentColumn\in\alphabet^\nSiteRows$, the domain $\alphabet^\nSiteRows$ of the mapping $\alignmentColumn \mapsto m_t(\alignmentColumn)$ is a finite set, and so all subsets of the domain are also finite.
This implies measurability of this mapping.
Finally, $m_t(\alignmentColumn)$ is nonpositive for all $t$ and $\alignmentColumn$, satisfying point four.
